# Simple Linear Regression {#sec-lm}

```{r status}
#| echo: false
#| output: show
source("_common.R")
display_chapter_status("drafting")
```

```{=latex}
\begin{objectives}{In this chapter, you will learn to}
\begin{itemize}

\item{Understanding the principles of simple linear regression}

\item{Performing simple linear regression in R for econometric analysis}

\item{Interpreting regression results in the context of economic variables}

\item{Assessing model assumptions and addressing violations}

\item{Practical examples and exercises using R}

\end{itemize}

\end{objectives}
```
```{r chapter-setup}
library(easystats)
library(ggplot2)
```

## Introduction

Linear regression is among the fundamental concepts in Econometrics. In linear regression, we try to estimate how much one variable will change, with response to a change in another variable. The controlled variable is called predictor or independent variable. The variable, which changes as a response to a change in the controlled variable is called response or dependent variable. In R, we denote such relationship with `y~x`, where y is the response and x is the predictor. The notation `y~x` is read as "y explained by x".

In mathematical terms, we are fitting a linear equation between x and y. When we write `y~x`, it means `y = ax + b`, and we need to find a and b from the data points given.

## Unveiling Relationships: The Purpose of Linear Regression

Linear regression, a cornerstone of statistical analysis, empowers us to delve into the fascinating world of relationships between variables. Imagine yourself as a data detective, piecing together clues to understand how one element influences another. This is precisely the role linear regression plays.

Suppose you're an enterprising researcher curious about the link between study hours and exam scores. Linear regression offers a method to uncover this connection. By harnessing the power of this technique, you can not only **predict** potential exam scores based on the number of hours a student studies, but also gain valuable insights into the **strength** of this relationship.

Consider the following scenario: you establish a model depicting the relationship between study hours and exam scores. This model, visualized as a straight line, allows you to predict the anticipated exam score for a student who dedicates a specific number of hours to studying. This predictive power is undeniably valuable.

However, linear regression goes beyond mere prediction. It unveils the **underlying nature** of the relationship between variables. The slope of the line you establish serves as a vital clue. A steeper slope suggests that additional study hours translate into a more significant increase in exam scores. Conversely, a gentler slope indicates a less pronounced impact of study hours on exam performance.

In essence, linear regression equips us with a two-pronged approach:

1.  **Prediction:** We can leverage the model to forecast outcomes based on new data points.

2.  **Understanding Relationships:** By analyzing the slope and other properties of the line, we gain insights into how changes in one variable (study hours) influence the other (exam scores).

It's important to remember that the world isn't always perfectly linear. However, linear regression serves as a robust foundation for data analysis, providing a framework to explore connections and make informed decisions based on the information gleaned from our data.

Let's understand what it means for us in terms of our workflow. The workflow for the linear regression problem would be:

1.  We would be given some observations of both - independent variable(s) and dependent variable.

2.  we graph these data points, using a coordinate system (like Cartesian system). Each value is represented by a dot. Such a diagram is called a scatter-plot diagram.

3.  After graphing the points onto a scatter plot diagram, linear regression analysis seeks to find the best-fit line to fit the points as closely as possible.

4.  This best-fit line is a line, which minimizes the distance between the points falling above or below the lines.

[Linear Regression 101: What Is It And How Is It Done? \| Fiverr](https://www.fiverr.com/resources/guides/data/linear-regression-101)

## Unveiling the Assumptions Behind Linear Regression

While linear regression offers a powerful lens for analyzing relationships, its effectiveness hinges on certain underlying assumptions. These assumptions, akin to the ground rules of the game, ensure the validity of the results we obtain. Let's delve into these key assumptions:

1.  **Linearity:** The core principle of linear regression lies in its name - **linearity**. This assumption dictates that the relationship between the two variables we're investigating is **straight-line** in nature. In simpler terms, when one variable increases or decreases, the other variable changes in a consistent, proportional manner, much like points scattered along a straight line.

2.  **Independence of Errors:** This assumption emphasizes that the **errors** (the difference between the actual data point and the value predicted by the model) are **independent** of each other. Imagine each error term as a random fluctuation; these fluctuations shouldn't be influenced by the errors associated with other data points.

3.  **Homoscedasticity:** This tongue-twister of a term refers to the notion of **constant variance** of the errors. In simpler terms, the spread of the errors (how far they deviate from the predicted value) should be **consistent** across all levels of the independent variable (the variable we believe causes the change in the other variable).

4.  **Normality of Errors:** This assumption posits that the errors associated with our data follow a **normal distribution**, often referred to as a bell curve. This characteristic distribution helps ensure the reliability of the statistical tests we perform to assess the model's validity.

By adhering to these assumptions, we create a solid foundation for interpreting the results of our linear regression analysis. However, it's crucial to remember that real-world data may not always perfectly satisfy these assumptions. Fortunately, there are techniques to address potential violations and ensure the robustness of our analysis. We'll explore these strategies in the upcoming chapters!

## Example in R

For understanding regression, we can generate a dummy dataset, and create a regression model on the basis of the data.

```{r dummy-data}
dummy <- data.frame(y = rnorm(100),x = rpois(100,3))
```

In the code above, we have created a dummy data. The `rnorm(n)` function generates n number of data points based on normal distribution. The default mean and standard deviation are 0 and 1 respectively. Similarly, the `rpois(n,l)` function generates n number of data points based on poisson distribution, with \lambda = l. We can plot the data to explore it visually. We will use `ggplot2` package to create this scatterplot diagram.

```{r}
#| label: fig-scatterplot
#| fig-cap: "Scatterplot of X and Y"
ggplot(dummy)+aes(x,y)+geom_point()+labs(title = "Scatterplot Diagram of X and Y")
```

Now we can create a regression model from this dataset.

```{r regression}
model1 <- lm(y~x, data = dummy)
```

This model stores a lot of information, but it is difficult to understand and get the meaning out of it. So, one way is to use the `summary` function on this `model1` object.

```{r regression-summary}
summary(model1)
```

The problem with this summary object is that it is not a data.frame. It is difficult to put this object in a research paper or any other academic submisison. So, we can use the `easystats` packages for this. The `model_parameters` function would show us the parameters (aka coefficients) of the model, and the `model_performance` function will show us the effectiveness of the regression model. We can also use the `display` function to beautify the table in the output.

```{r}
#| label: tbl-regression-params
#| tbl-cap: "Regression Parameters"
model1 |> model_parameters() |> 
  display(format = "markdown", caption = "Regression Parameters")
```

```{r}
#| label: tbl-regression-effectiveness
#| tbl-cap: "Regression Effectiveness"
model1 |> model_performance() |> 
  display(format = "markdown", caption = "Regression Effectiveness")
```

### Interpretation of Regression Model

\- The intercept is statistically non-significant and negative (beta = -0.16, 95% CI \[-0.59, 0.27\], t(98) = -0.74, p = 0.461; Std. beta = -5.50e-18, 95% CI \[-0.20, 0.20\]).

-   The effect of x is statistically non-significant and positive (beta = 0.07, 95% CI \[-0.04, 0.17\], t(98) = 1.24, p = 0.219; Std. beta = 0.12, 95% CI \[-0.07, 0.32\]).

The model explains a statistically not significant and very

weak proportion of variance (R\^2 = 0.02, F(1, 98) = 1.53, p =

0.219, adj. R\^2 = 5.32e-03).

### Non-Linear Relationship in Regression

Sometimes, the relationship between the dependent and independent variable is not linear, but quadratic. This can be understood either from the scatter plot diagram or from theoretical considerations. In mathematical terms, the relationship is

$$
Y = \beta_0 + \beta_1*X^2 + \gamma
$$ {#eq-quadratic}

where X and Y are independent and dependent variables, respectively. This is still considered "Linear Regression", because we are interested in the the linearity of the coefficient term (i.e., \beta), not the linearity of the variables. In the equation above, both the coefficients ($\beta_0$ and $\beta_1$) are linear, and so is the error term, $\gamma$ in this equation.

## Interval Estimation and Hypothesis Testing

Linear regression empowers us to understand relationships and make predictions. But remember, the world of data isn't perfect, and our models are based on samples, not the entire population. This inherent uncertainty is where **interval estimation** steps in.

Imagine you estimated a straight line to represent the connection between house size (square footage) and selling price. This line predicts the average selling price for a given house size. But what if a single house is much larger or smaller than those in your sample? The predicted price from the model might not be the exact selling price.

Interval estimation tackles this by providing a **range of plausible values** for the outcome variable (selling price) at a specific level of the independent variable (house size). This range, called a **confidence interval**, gives us a sense of the **precision** of our prediction.

Here's how it works:

1.  We choose a **confidence level**, typically 90% or 95%. This indicates the probability that the true value falls within the constructed interval.

2.  We leverage the concept of the **standard error of the estimate**. This statistical measure reflects the average amount our predictions might deviate from the actual values.

3.  By incorporating the chosen confidence level and the standard error, we calculate the **upper and lower bounds** of the confidence interval. This range represents the values within which we are confident (based on the chosen level) the true outcome variable will lie for a given independent variable value.

**What does this tell us?**

A narrower confidence interval indicates a more precise prediction. The model is more certain about the range of potential selling prices for a specific house size. Conversely, a wider interval reflects greater uncertainty. The model can't pinpoint the exact selling price as confidently.

**Interval estimation is crucial because:**

-   It acknowledges the inherent uncertainty in our predictions.

-   It allows us to communicate the **precision** of our findings.

-   It helps us assess the **reliability** of the model for predictions.

By incorporating interval estimation, we paint a more nuanced picture when interpreting the results of linear regression analysis. We not only provide a predicted value but also quantify the associated uncertainty, making our analysis more informative and reliable.

Let us understand interval estimation with an example of a real dataset. We will use the `mtcars` dataset, which comes bundled with R.

```{r}
mod1 <- lm(mpg~disp, data = mtcars)
```

In the code above, we fitted a model on the `mtcars` dataset, where Miles Per Gallon, or `mpg` is explained by Displacement (cubic inches), or `disp` variable. The \beta coefficient for `disp` is `r coef(mod1)[[2]]`. The 95% confidence interval of the coefficients can be found using the `confint` function in R.

```{r}
#| label: tbl-confint
#| tbl-cap: "Confidence Interval of Model Coefficients"
confint(mod1) |> kbl(digits = 4,booktabs = TRUE, align = "c")
```

In the code above, the `confint` function is used to find out the confidence interval, while the other functions are used to format the resulting table. The last row of the table suggests that the 95% confidence interval for the `disp` variable is between (-0.0508) and (-0.0316).

## Hypothesis Testing

Linear regression equips us to explore relationships and make predictions, but are these connections merely due to chance, or do they hold true significance? This is where **hypothesis testing** enters the scene. Imagine you suspect a link between coffee consumption and alertness levels. Regression analysis can tell you if there's a relationship, but hypothesis testing helps determine if this connection is statistically **meaningful**.

Here's the framework:

**Formulating Hypotheses:**

**Null Hypothesis (H₀):** This hypothesis proposes that there's **no significant relationship** between the independent and dependent variables (e.g., coffee consumption has no impact on alertness levels).

**Alternative Hypothesis (H₁):** This hypothesis stands in opposition to the null hypothesis, suggesting that a **significant relationship** exists (e.g., coffee consumption does influence alertness levels).

**Choosing a Significance Level:** This threshold, typically set at 5%, determines how likely we are to reject the null hypothesis even when it's actually true (a false positive). A lower significance level indicates a stricter test, requiring stronger evidence to reject the null hypothesis.

**Test Statistic and p-value:** We employ statistical tests like the t-test or F-test to calculate a **test statistic**. This value reflects the strength of the evidence against the null hypothesis. A higher (absolute) value suggests a stronger case for rejecting the null hypothesis.

We then calculate the **p-value**, which represents the probability of observing a test statistic as extreme or more extreme than the one we calculated, assuming the null hypothesis is true. A lower p-value indicates less likelihood of observing such an extreme result by chance alone.

**Decision Making:** Based on the chosen significance level and the p-value, we make a decision:

**Reject H₀:** If the p-value is lower than the significance level, we reject the null hypothesis. There's sufficient evidence to suggest a significant relationship between the variables.

**Fail to Reject H₀:** If the p-value is greater than the significance level, we fail to reject the null hypothesis. We cannot conclude a statistically significant relationship at this level of significance. However, this doesn't necessarily imply there's no connection; it just means we don't have enough evidence to disprove the null hypothesis.

R automatically performs a hypothesis testing, along with presenting regression summary. For example, in the @tbl-regression-params, the last column titled "p" conducts a two-way hypothesis testing for the t-value shown in the previous column, titled "t(98)", (where 98 number shows the degrees of freedom) whether the t-value is significantly different from zero. In this case, the null and the alternative hypothesis are as follows:

$$
H_0: t = 0
$$ $$
H_1: t \neq 0
$$

Given that both the p-values (0.234 and 0.324) are greater than the significance level (0.05), we fail to reject the null hypothesis. In other words, we believe that both, the intercept and x, are statistically significant in order to explain the variation in the dependent variable, y.

## Prediction

Linear regression not only unveils relationships between variables but also empowers us to **predict** future outcomes. Imagine you're a business analyst trying to forecast sales based on advertising spending. R, a powerful statistical programming language, provides tools to leverage your regression model for predictions.

Here's a breakdown of the prediction process with an example using R code:

**1. Sample Data and Model Fitting:**

```{r}
# Simulate some data for advertising spend (ad_spend) and sales (sales) 
set.seed(123) 
ad_spend <- rnorm(100, mean = 100, sd = 15) 
sales <- 50 + 2 * ad_spend + rnorm(100, mean = 0, sd = 10)  
# Fit the linear regression model
model <- lm(sales ~ ad_spend)
```

In this code:

-   We simulate data for advertising spend and sales.

-   The `lm` function fits a linear regression model where `sales` is the dependent variable and `ad_spend` is the independent variable.

**2. Predicting Sales for a New Advertising Spend Value:**

```{r}
# Predict sales for an ad spend of $120 
new_spend <- 120 
predicted_sales <- predict(model, newdata = data.frame(ad_spend = new_spend))  
cat("Predicted sales for an ad spend of $", new_spend, ":", predicted_sales, sep = "")
```

This code:

-   Defines a new advertising spend value of \$120.

-   The `predict` function leverages the fitted model (`model`) to predict sales for this new data point (represented as a data frame with the new `ad_spend` value).

-   The `cat` function displays the predicted sales value.

This indicates that based on the model, sales are predicted to be around \$290 for an advertising spend of \$120.

The key difference between prediction intervals and confidence intervals in the `predict` function of R lies in what they are trying to estimate:

-   **Prediction Interval:** Focuses on a **single future observation**. It estimates the range within which a **new individual value** of the dependent variable is likely to fall, given a specific value of the independent variable.

-   **Confidence Interval:** Focuses on the **mean prediction**. It estimates the range that's likely to contain the **average value** of the dependent variable for a specific value of the independent variable.

Here's a breakdown to illustrate the concept:

**Scenario:** Predicting sales based on advertising spend.

-   **Confidence Interval:** If you set a 95% confidence interval for predicted sales at an ad spend of \$120, it indicates that you're 95% confident the **average sales** for that advertising spend will fall within that interval.

-   **Prediction Interval:** A 95% prediction interval for the same scenario estimates the range that's likely to encompass a **single new sales value** at an ad spend of \$120.

**R Code Demonstration:**

```{r}
# Simulate data for advertising spend (ad_spend) and sales (sales)
set.seed(123) 
ad_spend <- rnorm(100, mean = 100, sd = 15) 
sales <- 50 + 2 * ad_spend + rnorm(100, mean = 0, sd = 10)  
# Fit the linear regression model
model <- lm(sales ~ ad_spend)  
# New advertising spend value
new_spend <- 120  
# Confidence Interval for Mean Prediction 
conf_interval <- predict(model, newdata = data.frame(ad_spend = new_spend), interval = "confidence")  
# Prediction Interval for a New Observation 
prediction_interval <- predict(model, newdata = data.frame(ad_spend = new_spend), interval = "prediction")  
cat("Predicted sales for an ad spend of $", new_spend, ":", summary(predicted_sales), sep = "\n") 
cat("95% Confidence Interval for Mean Prediction:", conf_interval[,c(1,3)], sep = "\n") 
cat("95% Prediction Interval for a New Observation:", prediction_interval[,c(1,3)], sep = "\n")
```

**Key Points:**

-   The confidence interval for the mean prediction is much narrower than the prediction interval for a new observation. This reflects the inherent uncertainty in predicting a single future value compared to the average sales across multiple observations.

-   Prediction intervals are generally wider because they account for the variability around the average sales figure, along with the uncertainty in the model itself.

In conclusion, choosing between a confidence interval and a prediction interval depends on your goal:

-   If you're interested in the **average sales** for a specific advertising spend, use the confidence interval.

-   If you want to estimate the range for a **single new sales value**, then the prediction interval is more appropriate.
